Nodal Discontinuous Galerkin Method for the Euler Equations of Fluid Dynamics

### 2017-01-09 Documents/Euler/dgEulerCurvilinear.tex
Adding transport application
Updates to hydro document

\title{Nodal Discontinuous Galerkin Method for the Euler Equations with Curvilinear Coordinates and Nuclear Equation of State for Astrophysics Simulations}
\author{Eirik Endeve\altaffilmark{1}}
\altaffiltext{1}{Department of Computational and Applied Mathematics, Oak Ridge National Laboratory, Oak Ridge, TN 37831-6354, USA; endevee@ornl.gov}

\begin{abstract}
We explore the use of the nodal Discontinuous Galerkin (DG) method in astrophysical hydrodynamics.  
We focus on issues related to core-collapse supernova dynamics, which include the use of curvilinear coordinates and a nuclear equation of state.  
\end{abstract}

@@

We consider evolution in three-dimensional position space and denote the position vector with for a point with coordinates $\bx=\{x^{1},x^{2},x^{3}\}$ with $\vect{r}(\vect{x})\in\mathbb{R}^{3}$, so that the displacement vector is $d\vect{r}=dx^{i}\,\vect{e}_{i}$ and the coordinate basis vectors are
\begin{equation}
  \vect{e}_{i}=\pderiv{\vect{r}}{x^{i}}.  
\end{equation}
The squared infinitesimal line-element is
\begin{equation}
  ds_{\vect{x}}^{2}=d\vect{r}\cdot d\vect{r}=\gamma_{ij}\,dx^{i}\,dx^{j},
\end{equation}
where the coordinate basis metric tensor is defined as $\gamma_{ij}=\vect{e}_{i}\cdot\vect{e}_{j}$.

For general curvilinear spatial coordinates, the Euler equations for gas dynamics are \citep{leveque_2002}
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\rho\,v^{i}\,\Big)=0,
\end{equation}
\begin{equation}
  \pderiv{M_{j}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,P^{i}_{~j}\,\Big)
  =\f{1}{2}\,P^{ik}\,\pderiv{\gamma_{ik}}{x^{j}}-\rho\,\pderiv{\Phi}{x^{j}},
\end{equation}
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\big(E+p\big)\,v^{i}\,\Big)
  =-M^{i}\,\pderiv{\Phi}{x^{i}},
\end{equation}
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,n_{e}\,v^{i}\,\Big)=0,
\end{equation}
where $\rho$ is the (baryon) mass density, $M^{i}=\rho\,v^{i}$ is the momentum density, $P^{ij}=\rho\,v^{i}\,v^{j}+p\,\gamma^{ij}$ the stress tensor, $E=e+\f{1}{2}\,\rho\,v_{i}\,v^{i}$ the total (internal plus kinetic) energy density, and $n_{e}$ the electron density.  
The pressure is related to the density, internal energy density, and electron density by an equation of state (EoS); i.e., $p=p(\rho,e,n_{e})$.  
(We also define the specific internal energy by $\epsilon=e/\rho$.)

We assume the following diagonal form of the spatial metric tensor
\begin{equation}
  \gamma_{ij}=\mbox{diag}\big[\,1,a(x^{1})^{2},\,b(x^{1})^{2}c(x^{2})^{2}\,\big], 
\end{equation}
which is sufficiently general to accommodate Cartesian, cylindrical, and spherical polar coordinates (see Table~\ref{tab:metricFunctions} and Appendix~\ref{app:CurvilinearEuler}).  
$\sqrt{\gamma}=abc$.  $w^{i}=e^{i}_{~j}\,\bar{w}^{j}$.  
With $e^{i}_{~j}=\mbox{diag}\big[\,1,a(x^{1})^{-1},\,b(x^{1})^{-1}c(x^{2})^{-1}\,\big]$, we have $\{w^{i}\}_{i=1}^{3}=\{\bar{w}^{1},\,a^{-1}\,\bar{w}^{2},\,b^{-1}c^{-1}\,\bar{w}^{3}\}$.  
\begin{table}
  \begin{center}
  \caption{Metric functions for Cartesian, Cylindrical, and Spherical coordinate systems.\label{tab:metricFunctions}}
  \begin{tabular}{cccccccccccccc}
    \midrule
    Coordinates & $x^{1}$ & $x^{2}$ & $x^{3}$ & $\bar{w}^{1}$ & $\bar{w}^{2}$ & $\bar{w}^{3}$ & $a$ & $b$ & $c$ & $\sqrt{\gamma}$
    & $\pderiv{a}{x^{1}}$ & $\pderiv{b}{x^{1}}$ & $\pderiv{c}{x^{2}}$ \\
    \midrule
    \midrule
    Cartesian & $x$ & $y$ & $z$ & $w_{x}$ & $w_{y}$ & $w_{z}$ & 1 & 1 & 1 & 1 & 0 & 0 & 0 \\
    Cylindrical & $R$ & $z$ & $\phi$ & $w_{R}$ & $w_{z}$ & $w_{\phi}$ & 1 & $R$ & 1 & $R$ & 0 & 1 & 0 \\
    Spherical & $r$ & $\theta$ & $\phi$ & $w_{r}$ & $w_{\theta}$ & $w_{\phi}$ & $r$ & $r$ & $\sin\theta$ & $r^{2}\sin\theta$ & 1 & 1 & $\cos\theta$ \\
    \midrule
    \midrule
  \end{tabular}
  \end{center}
\end{table}

The mass conservation equation
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\rho\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\rho\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\rho\,\bar{v}^{3}\,\Big)=0
\end{equation}
Components of the momentum equation
\begin{equation}
  \pderiv{\bar{M}^{i}}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\bar{P}^{1i}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\bar{P}^{2i}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\bar{P}^{3i}\,\Big)
  =\bar{S}^{i}+\bar{G}^{i},
\end{equation}
where the ``geometry sources'' are given by
\small
\begin{equation}
  \bar{S}^{1}
  =\bar{P}^{22}\f{1}{a}\pderiv{a}{x^{1}}+\bar{P}^{33}\f{1}{b}\pderiv{b}{x^{1}}, \quad
  \bar{S}^{2}
  =\bar{P}^{33}\f{1}{ac}\pderiv{c}{x^{2}}-\bar{P}^{12}\f{1}{a}\pderiv{a}{x^{1}}, \quad
  \bar{S}^{3}
  =-\bar{P}^{13}\f{1}{b}\pderiv{b}{x^{1}}-\bar{P}^{23}\f{1}{ac}\pderiv{c}{x^{2}},
\end{equation}
\normalsize
and the gravitational sources are
\begin{equation}
  \bar{G}^{1}
  =-\rho\,\pderiv{\Phi}{x^{1}}, \quad
  \bar{G}^{2}
  =-\rho\,\f{1}{a}\pderiv{\Phi}{x^{2}}, \quad
  \bar{G}^{3}
  =-\rho\,\f{1}{bc}\pderiv{\Phi}{x^{3}}.
\end{equation}
The energy equation is given by
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\big(E+p\big)\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\big(E+p\big)\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\big(E+p\big)\,\bar{v}^{3}\,\Big)=Q,  
\end{equation}
where the gravitational energy source is
\begin{equation}
  Q=-\Big(\,\bar{M}^{1}\pderiv{\Phi}{x^{1}}+\bar{M}^{2}\f{1}{a}\pderiv{\Phi}{x^{2}}+\bar{M}^{3}\f{1}{bc}\pderiv{\Phi}{x^{3}}\,\Big).
\end{equation}
Finally, the electron conservation equation
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,n_{e}\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,n_{e}\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,n_{e}\,\bar{v}^{3}\,\Big)=0
\end{equation}
By defining the vector of evolved variables $\bU=\big(\,\rho,\,\rho\,\bar{v}^{1},\,\rho\,\bar{v}^{2},\,\rho\,\bar{v}^{3},\,E,\,n_{e}\,\big)^{T}$, the vector of geometry sources $\vect{S}(\bU)=\big(\,0,\,\bar{S}^{1},\,\bar{S}^{2},\,\bar{S}^{3}\,,0,\,0\,\big)^{T}$, the vector of gravitational sources $\bG(\bU,\Phi)=\big(\,0,\bar{G}^{1},\bar{G}^{2},\bar{G}^{3},Q,0\,\big)^{T}$, and the vectors of flux vector components
\begin{equation}
  \vect{F}^{i}(\bU)
  =\big(\,\rho\,\bar{v}^{i},\,\rho\,\bar{v}^{1}\,\bar{v}^{i}+p\,\delta^{1i},\,\rho\,\bar{v}^{2}\,\bar{v}^{i}+p\,\delta^{2i},\,\rho\,\bar{v}^{3}\,\bar{v}^{i}+p\,\delta^{3i},\,(E+p)\,\bar{v}^{i},\,n_{e}\,\bar{v}^{i}\,\big)^{T},
\end{equation}
we can write the system of equations in compact form
\begin{equation}
  \pderiv{\bU}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\vect{F}^{1}(\bU)\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\vect{F}^{2}(\bU)\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\vect{F}^{3}(\bU)\,\Big)
  =\vect{S}(\bU)+\bG(\bU,\Phi).
  \label{eq:extendedEulerCompact}
\end{equation}

@@

@@

Inserting \eqref{eq:conservedNodalExpansion} into \eqref{eq:semidiscreteDG}, with $v(\bx)=\ell_{\bk}(\bx)$ and the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{\bK}\bU_{\DG}\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,|\bK|\,\pd{}{t}\bU_{\bk} \nonumber \\
  &\hspace{12pt}
  =w_{\bk}\,a(x_{k_{1}}^{1})\,b(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})\,\dx^{1}\,\dx^{2}\,\dx^{3}\,\pd{}{t}\bU_{\bk}
\end{align}
for the time derivative term, and
\begin{align}
  \int_{\bK}\bS(\bU_{\DG})\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,|\bK|\,\bS(\bU_{\bk}) \nonumber \\
  &\hspace{12pt}
  =w_{\bk}\,a(x_{k_{1}}^{1})\,b(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})\,\dx^{1}\,\dx^{2}\,\dx^{3}\,\bS(\bU_{\bk})
\end{align}
for the source term.  
For the surface integrals, we obtain, e.g., 
\begin{align}
  %
  \int_{\tilde{\bK}^{1}}v\,\widehat{\bF}^{1}|_{x_{\Hgh}^{1}}\,dA^{1}
  &\approx w_{\tilde{\bk}_{1}}\,\ell_{k_{1}}(+1)\,\widehat{\bF}^{1}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,
  \tau^{1}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,|\tilde{\bK}^{1}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{1}}\,\ell_{k_{1}}(+1)\,\widehat{\bF}^{1}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,
  a(x_{\Hgh}^{1})\,b(x_{\Hgh}^{1})\,c(x_{k_{2}}^{2})\,\dx^{2}\,\dx^{3}, \\
  %
  \int_{\tilde{\bK}^{2}}v\,\widehat{\bF}^{2}|_{x_{\Hgh}^{2}}\,dA^{2}
  &\approx w_{\tilde{\bk}_{2}}\,\ell_{k_{2}}(+1)\,\widehat{\bF}^{2}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,
  \tau^{2}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,|\tilde{\bK}^{2}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{2}}\,\ell_{k_{2}}(+1)\,\widehat{\bF}^{2}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,
  b(x_{k_{1}}^{1})\,c(x_{\Hgh}^{2})\,\dx^{1}\,\dx^{3}, \\
  %
  \int_{\tilde{\bK}^{3}}v\,\widehat{\bF}^{3}|_{x_{\Hgh}^{3}}\,dA^{3}
  &\approx w_{\tilde{\bk}_{3}}\,\ell_{k_{3}}(+1)\,\widehat{\bF}^{3}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,
  \tau^{3}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,|\tilde{\bK}^{3}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{3}}\,\ell_{k_{3}}(+1)\,\widehat{\bF}^{3}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,
  a(x_{k_{1}}^{1})\,\dx^{1}\,\dx^{2}.  
\end{align}
Finally, the `volume terms' become
\begin{align}
  %
  \int_{\bK}\bF^{1}\,\pderiv{v}{x^{1}}\,dV
  &\approx w_{\tilde{\bk}_{1}}\sum_{q_{1}=1}^{N}w_{q_{1}}\,\bF^{1}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,
  \pderiv{\ell_{k_{1}}}{\eta^{1}}(\eta_{q_{1}}^{1})\,\sqrt{\gamma}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,|\tilde{\bK}^{1}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{1}}\,c(x_{k_{2}}^{2})\,\dx^{2}\,\dx^{3}
  \sum_{q_{1}=1}^{N}w_{q_{1}}\,\bF^{1}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,
  \pderiv{\ell_{k_{1}}}{\eta^{1}}(\eta_{q_{1}}^{1})\,a(x_{q_{1}}^{1})\,b(x_{q_{1}}^{1}), \\
  %
  \int_{\bK}\bF^{2}\,\f{1}{a}\pderiv{v}{x^{2}}\,dV
  &\approx w_{\tilde{\bk}_{2}}\sum_{q_{2}=1}^{N}w_{q_{2}}\,\bF^{2}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,
  \f{1}{a(x_{k_{1}}^{1})}\pderiv{\ell_{k_{2}}}{\eta^{2}}(\eta_{q_{2}}^{2})\,\sqrt{\gamma}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,|\tilde{\bK}^{2}| \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{2}}\,b(x_{k_{1}}^{1})\,\dx^{1}\,\dx^{3}\sum_{q_{2}=1}^{N}w_{q_{2}}\,\bF^{2}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,
  \pderiv{\ell_{k_{2}}}{\eta^{2}}(\eta_{q_{2}}^{2})\,c(x_{q_{2}}^{2}), \\
  %
  \int_{\bK}\bF^{3}\,\f{1}{bc}\pderiv{v}{x^{3}}\,dV
  &\approx w_{\tilde{\bk}_{3}}\sum_{q_{3}=1}^{N}w_{q_{3}}\,\bF^{3}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,
  \f{1}{b(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})}\pderiv{\ell_{k_{3}}}{\eta^{3}}(\eta_{q_{3}}^{3})\,\sqrt{\gamma}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,|\tilde{\bK}^{3}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{3}}\,a(x_{k_{1}}^{1})\,\dx^{1}\,\dx^{2}\sum_{q_{3}=1}^{N}w_{q_{3}}\,\bF^{3}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,
  \pderiv{\ell_{k_{3}}}{\eta^{3}}(\eta_{q_{3}}^{3}).  
\end{align}

>>>>

\title{Nodal Discontinuous Galerkin Method for the Euler Equations of Fluid Dynamics}
\author{Eirik Endeve\altaffilmark{1}, et al.}
\altaffiltext{1}{Computational and Applied Mathematics Group, Oak Ridge National Laboratory, Oak Ridge, TN 37831-6354, USA; endevee@ornl.gov}

\begin{abstract}
We present a nodal Discontinuous Galerkin (DG) method for the Euler equations with an eye towards astrophysical hydrodynamics.  
We focus on issues related to core-collapse supernova dynamics, which include the use of curvilinear coordinates and a non-ideal (nuclear) equation of state.  
\end{abstract}

\tableofcontents

@@

\subsection{Basic Mathematical Model}

For general curvilinear spatial coordinates, the non-relativistic Euler equations for gas dynamics are \citep[][see, e.g., \citet{} and references therein for a fully general relativistic treatment]{leveque_2002}
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\rho\,v^{i}\,\Big)=0,
  \label{eq:massConservation}
\end{equation}
\begin{equation}
  \pderiv{S_{j}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,P^{i}_{~j}\,\Big)
  =\f{1}{2}\,P^{ik}\,\pderiv{\gamma_{ik}}{x^{j}}-\rho\,\pderiv{\Phi}{x^{j}},
  \label{eq:momentumConservation}
\end{equation}
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\big(E+p\big)\,v^{i}\,\Big)
  =-S^{i}\,\pderiv{\Phi}{x^{i}},
  \label{eq:energyConservation}
\end{equation}
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,n_{e}\,v^{i}\,\Big)=0,
  \label{eq:electronConservation}
\end{equation}
where $\rho$ is the (baryon) mass density, $S^{i}=\rho\,v^{i}$ is the momentum density, $P^{ij}=\rho\,v^{i}\,v^{j}+p\,\gamma^{ij}$ the stress tensor, $E=e+\f{1}{2}\,\rho\,v_{i}\,v^{i}$ the total (internal plus kinetic) energy density, and $n_{e}$ the electron density.  
The pressure is related to the density, internal energy density, and electron density by an equation of state (EoS); i.e., $p=p(\rho,e,n_{e})$.  
(We also define the specific internal energy by $\epsilon=e/\rho$.)
The Newtonian gravitation potential, $\Phi$, is obtained from solving
\begin{equation}
  \f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\gamma^{ij}\pderiv{\Phi}{x^{j}}\,\Big)=4\pi\,\rho.  
  \label{eq:poisson}
\end{equation}

The model given by Eqs.~\eqref{eq:massConservation} - \eqref{eq:electronConservation} is non-relativistic (i.e., it does not include effects due to strong gravitational fields or relativistic fluid velocities), but is expressed in terms of curvilinear coordinates through the spatial metric tensor $\gamma_{ij}$, which gives squared infinitesimal line-element
\begin{equation}
  ds_{\vect{x}}^{2}=\gamma_{ij}\,dx^{i}\,dx^{j},
\end{equation}
where $\gamma_{ij}$ are the spatial components of the coordinate basis metric tensor.  
We will make two simplifying assumptions: the metric tensor is (1) diagonal and (2) time-independent.  
Specifically we assume the following diagonal form
\begin{equation}
  \gamma_{ij}
  =\mbox{diag}\big[\,\gamma_{11},\,\gamma_{22}(x^{1}),\,\gamma_{33}(x^{1},x^{2})\,\big]
  =\mbox{diag}\big[\,1,\,a(x^{1})^{2},\,b(x^{1})^{2}c(x^{2})^{2}\,\big], 
  \label{eq:threeMetric}
\end{equation}
which is sufficiently general to accommodate Cartesian, cylindrical, and spherical polar coordinates (see Table~\ref{tab:metricFunctions} and Appendix~\ref{app:CurvilinearEuler}).  
Moreover, $\gamma$ is the determinant of the metric tensor, and $\sqrt{\gamma}=abc$.  
The inverse of the spatial metric is denoted $\gamma^{ij}$, so that $\gamma^{ik}\gamma_{kj}=\delta^{i}_{~j}$.  
We can use the spatial metric to raise and lower indices on spatial vectors and tensors; e.g., $S_{j}=\gamma_{jk}\,S^{k}$.

\begin{table}
  \begin{center}
  \caption{Metric functions for Cartesian, Cylindrical, and Spherical coordinate systems.\label{tab:metricFunctions}}
  \begin{tabular}{cccccccccccccc}
    \midrule
    Coordinates & $x^{1}$ & $x^{2}$ & $x^{3}$ & $\bar{w}^{1}$ & $\bar{w}^{2}$ & $\bar{w}^{3}$ & $a$ & $b$ & $c$ & $\sqrt{\gamma}$
    & $\pderiv{a}{x^{1}}$ & $\pderiv{b}{x^{1}}$ & $\pderiv{c}{x^{2}}$ \\
    \midrule
    \midrule
    Cartesian & $x$ & $y$ & $z$ & $w_{x}$ & $w_{y}$ & $w_{z}$ & 1 & 1 & 1 & 1 & 0 & 0 & 0 \\
    Cylindrical & $R$ & $z$ & $\phi$ & $w_{R}$ & $w_{z}$ & $w_{\phi}$ & 1 & $R$ & 1 & $R$ & 0 & 1 & 0 \\
    Spherical & $r$ & $\theta$ & $\phi$ & $w_{r}$ & $w_{\theta}$ & $w_{\phi}$ & $r$ & $r$ & $\sin\theta$ & $r^{2}\sin\theta$ & 1 & 1 & $\cos\theta$ \\
    \midrule
    \midrule
  \end{tabular}
  \end{center}
\end{table}

\subsection{Euler Equations as a System of Conservation Laws with Sources}

We define the geometry sources (fictitious plus gravitational sources) in Eq.~\eqref{eq:momentumConservation} by
\begin{equation}
  G_{j}
  =\f{1}{2}\,P^{ik}\,\pderiv{\gamma_{ik}}{x^{j}}-\rho\,\pderiv{\Phi}{x^{j}}.  
\end{equation}
Specifically, using the time-independent metric tensor in Eq.~\eqref{eq:threeMetric}, the components are
\begin{equation}
  G_{1}
  =\f{1}{2}\,\Big(\,P^{22}\,\pderiv{\gamma_{22}}{x^{1}}+P^{33}\,\pderiv{\gamma_{33}}{x^{1}}\,\Big)-\rho\,\pderiv{\Phi}{x^{1}}, \quad
  G_{2}
  =\f{1}{2}\,P^{33}\,\pderiv{\gamma_{33}}{x^{2}}-\rho\,\pderiv{\Phi}{x^{2}}, \quad\text{and}\quad
  G_{3}
  =-\rho\,\pderiv{\Phi}{x^{3}}.
\end{equation}
The gravitational energy source term in Eq.~\eqref{eq:energyConservation} is denoted with $G_{E}=-S^{i}\,\pd{\Phi}{x^{i}}$.

Then, by defining the vector of evolved variables $\bU=\big(\,\rho,\,S_{1},\,S_{2},\,S_{3},\,E,\,n_{e}\,\big)^{T}$, the vector of geometry sources $\bG(\bU)=\big(\,0,\,G_{1},\,G_{2},\,G_{3},\,G_{E},\,0\,\big)^{T}$, and the flux vector components
\begin{equation}
  \vect{F}^{i}(\bU)
  =\big(\,\rho\,v^{i},\,P_{~1}^{i},\,P_{~2}^{i},\,P_{~3}^{i},\,(E+p)\,\bar{v}^{i},\,n_{e}\,\bar{v}^{i}\,\big)^{T},
\end{equation}
we write the system of equations in the usual compact form
\begin{equation}
  \pd{\bU}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\vect{F}^{i}(\bU)\,\Big)
  =\bG(\bU).
  \label{eq:extendedEulerCompact}
\end{equation}
Eq.~\eqref{eq:extendedEulerCompact} forms the basis for developing DG methods for the Euler equations.

@@

\subsection{Basic Principles of the DG Method}

@@

Inserting \eqref{eq:conservedNodalExpansion} into \eqref{eq:semidiscreteDG}, with $v(\bx)=\ell_{\bk}(\bx)$ and the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{\bK}\bU_{\DG}\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,\pd{}{t}\bU_{\bk}\,|\bK|
\end{align}
for the time derivative term, and
\begin{align}
  \int_{\bK}\bG(\bU_{\DG})\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,\bG(\bU_{\bk})\,|\bK|.
\end{align}
for the source term.  
The gravitational source terms contain the derivative to the potential, which warrants some additional exposition.  
We choose a local representation of the potential (obtained by evaluating the potential in the nodes in $\bK$ using the FEM solution); i.e.,
\begin{equation}
  \Phi(\bx)=\sum_{\bi=\vect{1}}^{\bN}\Phi_{\bi}\,\ell_{\bi}(\bx).  
\end{equation}
Differentiating with respect to $x^{j}$ gives
\begin{equation}
  \pderiv{\Phi}{x^{j}}\Big|_{\bx_{\bk}}
  =\f{1}{|K^{j}|}\sum_{\bi=\vect{1}}^{\bN}\pderiv{\ell_{\bi}}{\eta^{j}}(\eta_{\bk})\,\Phi_{\bi}
  =\f{1}{|K^{j}|}\sum_{i_{j}=1}^{N}\pderiv{\ell_{i_{j}}}{\eta^{j}}(\eta_{k_{j}}^{j})\,\Phi(x_{i_{j}}^{j},\tilde{\bx}_{\tilde{\bk}_{j}}^{j}).  
\end{equation}
This expression is used to evaluate the gravitational sources in the momentum and energy equations.
